The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-23 manuscript/rebuttal.Rmd

Response: We appreciate reviewer 2's positive feedback, insights and thoughtful suggestions.

### R2 Point 1

> 1. Inclusion of Case Studies or Examples: To strengthen your argument, consider incorporating case
> studies or practical examples that demonstrate the effectiveness of your recommendations.
> Highlighting specific instances where these best practices have enhanced reproducibility or solved
> particular issues could substantiate your points more effectively.

RESPONSE: To demonstrate the importance of computational reprodcibility, we elaborate on the
"forensic bioinformatics" case study of Baggerly and Coombes (2010) critique of Potti et al (2006)
in the introduction:

@@

> `git` is typically used at the command line, however it is also incorporated into integrated
> development environments commonly used in bioinformatics including Rstudio, JupyterLab [68,69] and
> VS Code [70,71].

### R3 Point 15

> 15. Page 9, line 7: References are missing for these other tools.

RESPONSE: Citations have been added for Podman and Apptainer/Singularity.
Rkt was removed, as it appears that development of the project has ceased.

### R3 Point 16

@@

> The [re3data.org](https://www.re3data.org/) registry of data repositories may be useful to find
> FAIR repositories that accepts data from other domains of study like ecology,
> physiology, molecular simulation, social sciences and computing [@Pampel2013-ep].

### R3 Point 23

> 23. Page 11, line 34: Need evidence to back this up.

RESPONSE: Reviewer 3 commented on the passage saying supplementary data files are less
findable and accessible as compared to FAIR data repositories.
We note that Roche and colleagues previously stated that [@Roche2015-aq]:

> There are no standards for organizing supplementary data both within and across journals
> and such data are often not readily discoverable or openly accessible
> (to those without a relevant journal subscription, for example).

They also cite an article by Caetano and Aisenberg that speaks directly to the findability and
accessibility @Caetano2014-bi:

> Many journals provide the option for authors to
> include supplementary data that are made available in the digital version of the publication.
> However, repositories increase data discoverability, are more reliable for long-term storage and
> make it possible to make data sets available under open access even
> when authors transfer copyright ownership to publishers.

In light of this argument, we have added a citation to Caetano and Aisenberg in the updated text.

### R3 Point 24

@@

> 24. Page 11, line 39: It is not necessarily true that CSV files are better than Excel.
Excel can retain information about data types, for example, whereas CSVs do not.
It depends on what you are trying to accomplish.

RESPONSE: It is true that in some cases, Excel files are more convenient to view.
However, in the context of large data sets with thousands of rows and their analytical
reproducibility, Excel is a non-starter.
For example, it is well known that the behaviour of Excel changes between versions.
Earlier versions have limitations around the number of allowed rows and columns.
Small changes to Excel files such as merging cells can break machine readibility.
Excel users often use cell highlighting or text colour to encode some information which isn't
easily machine readable.
The simplicity of TSV and CSV formats means they are readily and rapidly ingested by commonly used
programming languages without needing to install extra packages.
In line with this, the point in question has been amended:

> 5. Use file formats that are machine-readable and compatible with many different types of software.
> For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
> specifications that are suitable for very large data.

### R3 Point 25

@@

> 25. One thing that could be added is something about Common Workflow Language.
It's a community-supported specification for accomplishing many of the objectives described here.
There are some recent papers about this.

TODO:

### R3 Point 26

> 26. Page 14, lines 9-10: The big question is how to make more progress.
> We have the tools to achieve reproducibility, but why are we rarely achieving it?
> The paper mentions incentives and lack of training, which are true.
> You might consider elaborating a bit.
> My cynical view is that writing more papers and tutorials will do little without strong incentives
> and more automation.

### 2023-08-23 manuscript/5pillars.Rmd
rev3 nearly done

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman, Rkt or Apptainer/Singularity, Docker remains the most widely used
containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

3. Avoid large supplementary data files accompanying journal articles as these are less findable and
accessible.

@@

5. Use file formats that are compatible with many different types of software; e.g.: comma-separated
values (CSV) files are preferred over Excel files.

>>>>

Containers are an attempt to resolve some of the downsides of VMs.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the five containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman [@Heon2018-lc] or Apptainer/Singularity
[@Kurtzer2017-zq], Docker remains the most widely used containerisation system for web development and research.
Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images in the DockerHub registry.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei; @Mangul2019-xp; @Krafczyk2021-xv].

@@

3. Avoid large supplementary data files accompanying journal articles as these are less findable and
accessible [@Caetano2014-bi].

@@

5. Use file formats that are machine-readable and compatible with many different types of software.
For example, the comma-separated values (CSV) and tab-separated values (TSV) formats are simple file
specifications that are suitable for very large data.
